Supplementary Information

### 2020-12-21 doc/manuscript/MEE_manuscript.Rmd
updating results

```{r sites, eval=FALSE, echo=FALSE}
# This generates a table, but doesn't format nicely and messes up special characters in names, so turned off.
library(knitr)  
#kableExtra  library will mess it up!
sites <- read.csv("tables_figures/sites.csv", stringsAsFactors = FALSE, check.names = FALSE)

knitr::kable(sites[, c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. Precip.",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)")], 
                  col.names = c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. PPT",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)"), 
                  caption = "Table 1. Sites included in this analysis.") 
```

@@

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records (Appendix `r climate_correction`). 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation ($PPT$); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration ($PET$) and precipitation day frequency ($PDF$).
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, *v1.0-alpha* [@anderson-teixeira_forestgeoclimate_2020].

@@

To test for $DBH \times$ climate interactions and $year$ effects, we limited the analysis of each species to years for which the $DBH$ range represented by at least 3 individuals spanned at least half the total range of that species, dropping species entirely if < 30 years met this criteria.  **(HAVE YET TO TRY THIS) **
For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.
For models including $year$, we tested for first-order linear effects of $year$.
To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we separately analyzed $DBH$-growth relationships by decade (Appendix `r decadal_analysis_appendix`). 
For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across metrics.

@@

**Using $RW$ as the growth metric, our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).**
While one-to-one correspondence of estimated climate coefficients was neither expected nor observed, estimates were correlated and rarely differed significantly (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).

@@

**The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$), consistently exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs. `r all_climwin`).**
In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
In # cases, *climwin* identified the same climate variable but different (sometimes overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
Finally, in *2 cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at BCI, precipitation variable group at HF; Figs. `r clim_sens`, **S5, S11**)*, *climwin* identified different climate variables and different time windows (e.g., Fig. `r climwin_diffV_diffW`). 
Henceforth, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)

Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*5 of 8* sites with both variables available), but was sometimes surpassed by precipitation frequency ($PDF$; n=*3 of 8* sites with both variables available). 
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver *at five and four of the ten sites, respectively*, noting that $PET$ was not available for two of these sites. 
$T_{min}$ was the top driver at only *one site (BCI)*, where it's effects were only marginally significant, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig. `r clim_sens`).

Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation. 
At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
Optimal windows for temperature variables tended to be shorter, the longest being a *6*-month period during the summer (wet season) at HKK. 
At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.
There were also a few instances where previous growing season conditions had the strongest influence.

>>>>

@@

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records (Appendix `r climate_correction`). 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation ($PPT$); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration ($PET$) and precipitation day frequency ($PDF$).
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeoclimate_2020].

@@

For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.

To test for $year$ effects, we limited the analysis of each species to years for which the $DBH$ range represented by at least 3 individuals spanned at least half the total range of that species, dropping species entirely if < 30 years met this criteria. 
We then ran models as described above, including a  first-order linear effect of $year$.
To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we separately analyzed $DBH$-growth relationships by decade (Appendix `r decadal_analysis_appendix`). 
For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across metrics.

@@

Using $RW$ as the growth metric, our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
While one-to-one correspondence of estimated climate coefficients was neither expected nor observed, estimates were correlated and rarely differed significantly (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).

@@

```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Dec.21)
#sameV_sameW: LDWp, NMt, NIOt, NIOp, SCBIp, SCt, ZOFp
#sameV_diffW: BCIp, BCIt, CBp, CBt, HFt, HKKp, LDWt, NMp, SCBIt, SCp, ZOFt
#diffV_diffW: HFp, HKKt
```
The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$), consistently exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs. `r all_climwin`).
In some cases (n= 7 of 20), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
In 11 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
Finally, two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs. `r clim_sens`, `r GLS_HKK`, `r GLS_HF`), *climwin* identified different climate variables and different time windows (e.g., Fig. `r climwin_diffV_diffW`). 
Henceforth, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed.)

```{r , eval=FALSE, echo=FALSE}
## Tallies for the above paragraph (Dec.21)
#PPT over PDF: 8 sites with both. PDF selected in HKK, HF, ZOF, Niobrara
#T group:Tmax at 5, PET at 3, Tmin at 1, Tmp at 1
```
Within the precipitation group, precipitation amount ($PPT$) and frequency ($PDF$) were each identified as the top variable in four of the eight sites with both variables available (Fig. `r clim_sens`). 
At the one site for which streamflow data were available (NIO), $PDF$ was selected as a stronger--*albiet* weak--driver (Figs. `r clim_sens`).
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two of these sites. 
$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCI, respectively), noting that the effects of $T_{min}$ at BCI were only marginally significant (Fig. `r clim_sens`).

```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Dec.21)
# BCI: precip dry season (12mo), T dry season (2 mo) 
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
#LDW: precip c June (1 mo), PET current June (1mo)
#HF: PDF pMay (1mo), PET cMar (1mo)
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
#NIO: wet cMay (1mo), TMP pJune (1mo)
#LT: pNov-cJul (9mo), TMX cMay-Jul (3mo)
#CB: cApr-cJul(4mo), TMXpJul-cJul(13mo)
#SC:pJun-pAug(3mo), TMXcJul (1mo)
```
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`): wet seasons in the tropics (precipitation variables at both BCI and HKK; $T_{max}$ at HKK) or late spring/ early summer in the extratropics (n= 5 of 8 each for temperature and precipitation variables).
In the tropics, the windows over which precipitation was influential included part (BCI) or all (HKK) of the dry season.
In the extratropics, there were two cases where temperature variables were most influential in late winter/early spring (March $PET$ at HF; Jan-March $T_{max}$ at ZOF; Fig. `r clim_sens`).
*(Results before this are up-to-date as of Dec. 21. The following need updating.)*
There were also cases where previous growing season conditions were influential:...
At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation. 
At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
Optimal windows for temperature variables tended to be shorter, the longest being a *6*-month period during the summer (wet season) at HKK. 
At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.
There were also a few instances where previous growing season conditions had the strongest influence.
